Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-08 talk/icooolps2009/paper.tex
some comments/fixes from toon

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand{\commentout}[1]{}

@@

During tracing, the trace is repeatedly
checked as to whether the interpreter is at a position in the program where it had been
earlier. If this happens, the trace recorded corresponds to a loop
in the interpreted program. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be executed immediately,
starting from the next iteration of the loop, as the machine code represents
exactly the loop that was being interpreted so far.

@@

Let's look at a small example. Take the following (slightly contrived) RPython
code:
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result
\end{verbatim}
}
To trace this, a bytecode form of these functions needs to be introduced that
the tracer understands. The tracer interprets a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed.
When the profiler shows
that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
tracing JIT will start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
The operations in this sequence are operations of the above-mentioned intermediate
representation (e.g. note that the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} is inlined into the trace. Note that the trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be converted into machine code and executed.

@@

Let's look at how hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. To apply hints one generally needs a
subclass of \texttt{JitDriver} that lists all the variables of the bytecode
loop. The variables are classified into two groups, red variables and green
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter.
However, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red.

@@

The simple insight how to improve the situation is that most of the
operations in the trace are actually concerned with manipulating the
bytecode and the program counter. Those are stored in variables that are part of
the position key (they are ``green''), that means that the tracer checks that they
are some fixed value at the beginning of the loop (they may well change over the
course of the loop, though). In the example the check
would be that the \texttt{bytecode} variable is the bytecode string
corresponding to the square function and that the \texttt{pc} variable is
\texttt{4}. Therefore it is possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} are additions of the green variable \texttt{pc} and a constant
number, so they can be folded away as well.

With this optimization enabled, the trace looks as in Figure
\ref{fig:trace-full}. Now much of the language interpreter is actually gone
from the trace and what is left corresponds very closely to the loop of the
square function. The only vestige of the language interpreter is the fact that
the register list is still used to store the state of the computation. This
could be removed by some other optimization, but is maybe not really all that
bad anyway (in fact we have an experimental optimization that does exactly that,
but it is not yet finished).  Once we get this optimized trace, we can pass it to
the \emph{JIT backend}, which generates the correspondent machine code.

@@

The first integration problem is how to \emph{not} integrate the tracing JIT at
all. It is possible to choose when the language interpreter is translated to C
whether the JIT should be built in or not. If the JIT is not enabled, all the
hints that are possibly in the interpreter source are just ignored by the
translation process. In this way, the result of the translation is identical to
that when no hints were present in the interpreter at all.

@@

\textbf{Benchmark 1:} The interpreter translated to C without any JIT inserted at all.

>>>>

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\toon[1]{\nb{TOON}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand{\commentout}[1]{}

@@

\toon{that part seems a bit out of place, it jumps back to a point in the story
before what you explained right before it. (as far as I can tell)}
During tracing, the trace is repeatedly
checked as to whether the interpreter is at a position in the program where it had been
earlier. If this happens, the trace recorded corresponds to a loop
in the interpreted program. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be executed immediately,
starting from the next iteration of the loop, as the machine code represents
exactly the loop that was being interpreted so far.

@@

Let's look at a small example. Take the following (slightly contrived) RPython
code:
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result
\end{verbatim}
}
\toon{next sentence is strange} To trace this, a bytecode form of these functions needs to be introduced that
the tracer understands. The tracer interprets a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed.
When the profiler discovers
that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
tracing JIT will start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
The operations in this sequence are operations of the above-mentioned intermediate
representation (e.g. the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} is inlined into the trace. The trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be converted into machine code and executed.

@@

Let's look at how hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. To apply hints one needs a
subclass of \texttt{JitDriver} that lists all the variables of the bytecode
loop. The variables are classified into two groups, red variables and green
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter.
However, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red (the fact
that red variables need to be listed explicitly too is an implementation
detail).

@@

\toon{very difficult to read (actually so is the whole paragraph; rephrase)}
The simple insight how to improve the situation is that most of the
operations in the trace are actually concerned with manipulating the
bytecode and the program counter. Those are stored in variables that are part of
the position key (they are ``green''), that means that the tracer checks that they
are some fixed value at the beginning of the loop (they may well change over the
course of the loop, though). In the example the check
would be that the \texttt{bytecode} variable is the bytecode string
corresponding to the square function and that the \texttt{pc} variable is
\texttt{4}. Therefore it is possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} are additions of the green variable \texttt{pc} and a constant
number, so they can be folded away as well.

With this optimization enabled, the trace looks as in Figure
\ref{fig:trace-full}. Now much of the language interpreter is actually gone
from the trace and what is left corresponds very closely to the loop of the
square function. The only vestige of the language interpreter is the fact that
the register list is still used to store the state of the computation. This
could be removed by some other optimization, but is maybe not really all that
bad anyway (in fact we have an experimental optimization that does exactly that,
but it is not yet finished).  Once we get this optimized trace, we can pass it to
the \emph{JIT backend}, which generates the corresponding machine code.

@@

The first integration problem is how to \emph{not} integrate the tracing JIT at
all. It is possible to choose when the language interpreter is translated to C
whether the JIT should be built in or not. If the JIT is not enabled, all the
hints that are possibly in the interpreter source are just ignored by the
translation process. In this way, the result of the translation is identical to
that when no hints were present in the interpreter at all. \toon{strange
sentence}

@@

\textbf{Benchmark 1:} The interpreter translated to C without the JIT.
